---
title: "ATiG 2026: PCA Exercise using bedpca — worked answers"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This exercise generally follows the vignette by Florian Privé available online [**here**](https://privefl.github.io/bigsnpr/articles/bedpca.html).

The most relevant literature is [Privé *et al.*, Bioinformatics 2020](https://doi.org/10.1093/bioinformatics/btaa520). And a draft manual for `bigstatsr` and `bigsnpr` can be found [here](https://privefl.github.io/bigsnpr-extdoc/introduction.html). 

## Prerequisites (IMPORTANT)

Use a current R installation and RStudio. Open this file and choose **Knit to HTML**. When running chunks interactively, set the working directory to this file's folder (RStudio: **Session > Set Working Directory > To Source File Location**).

Install the following packages once:

```{r install, eval=FALSE}
install.packages(c("bigsnpr", "ggplot2", "cowplot", "knitr", "rmarkdown"))
```

The first run downloads about 1 GB of genotype data plus PLINK 2; later runs reuse the files in `data/`. The analysis reads genotypes from disk and uses two cores by default.

```{r setup, include=FALSE}
knitr::opts_chunk$set(error = FALSE, fig.width = 8, fig.height = 5, dpi = 120)
set.seed(2026)
ncores <- as.integer(Sys.getenv("ATIG_PCA_NCORES", "2"))
stopifnot(length(ncores) == 1L, !is.na(ncores), ncores >= 1L)
```

First we load the packages.

```{r packages, message=FALSE}
library(bigsnpr)
library(ggplot2)
```

### 1 Principal Component Analysis (PCA) and Relatedness

A vignette for how to conduct PCA analyses using the package can be found [**here**](https://privefl.github.io/bigsnpr/articles/bedpca.html).

First we download the 1000 genomes data in plink bed file format, which can take time if the internet connection is slow.

```{r data}
dir.create("data", showWarnings = FALSE)
bedfile <- download_1000G("data")
```

##### Q1: What are plink bed files?

A bed file is one of the three primary PLINK files: .bed, .bim, and .fam, used to store and manipulate large-scale genotype data.
The bed file stores the actual genotype (0, 1, 2) information in a highly efficient binary format. The bim file stores chromosome, variant ID, genetic position, physical position, and alleles A1/A2. These are not guaranteed to be reference/alternative alleles; see the [PLINK file-format reference](https://www.cog-genomics.org/plink/1.9/formats#bim). Finally, the fam file stores information about the samples (individuals) in the study (family ID, individual ID, paternal ID, maternal ID, sex, phenotype (control/case/missing 1/2/-9)). Fam ID is often 0 or the same as individual ID, paternal and maternal IDs or often 0, because they're unknown.


#### Relatedness

First, let us detect all pairs of related individuals by filtering on the kinship coefficient. Note that the kinship coefficient is the probability that a pair of randomly sampled homologous alleles are identical by descent (IBD). In pedigree terms, the coefficient of relatedness is twice the kinship coefficient. An estimated genomic relationship matrix depends on its allele-frequency reference and is not automatically twice a KING estimate.


```{r relatedness}
plink2 <- download_plink2("data",
                          ARM = grepl("aarch64|arm64", R.version$arch),
                          AVX2 = FALSE)
# Keep relative paths: the course folder itself may contain spaces.
plink2 <- file.path("data", basename(plink2))
rel <- snp_plinkKINGQC(
  plink2.path = plink2,
  bedfile.in = bedfile,
  thr.king = 2^-4.5,
  make.bed = FALSE,
  ncores = ncores
)
str(rel)
```

##### Q2: What threshold was used to filter related individuals? What level of relatedness does it correspond to?

The threshold used to filter related individuals is 2^-4.5 (approx. 0.0442). 
This threshold corresponds to filtering out third-degree relationships or closer (e.g., great-grandparents). See `?snp_plinkKINGQC`

0.5 corresponds to duplicates or monozygotic twins.

~0.25 corresponds to first-degree relations (e.g., parent-child, full siblings).

~0.125 corresponds to second-degree relations (e.g., half-siblings, grandparents, aunts/uncles).

~0.0625 corresponds to third-degree relations (e.g., great-grandparents)

~0.03125 corresponds to fourth-degree relations (e.g., great-great-grandparents)


##### Q3: Can you plot the kinship values, also using a less stringent threshold?

```{r kinship-plot, fig.cap="Figure 1. Reported kinship estimates and the two filtering thresholds."}
hist(rel$KINSHIP, xlab = "KING kinship", main = "Reported related pairs",
     xlim = range(c(rel$KINSHIP, 2^-4.5, 2^-3.5)))
abline(v = c(2^-4.5, 2^-3.5), col = c("red", "darkgreen"))
```

Anything to the right of the red line would be considered too closely related and would be filtered out when using the 2^-4.5 threshold. Meanwhile, anything to the right of the green line would be considered too closely related and would be filtered out when using the less stringent 2^-3.5 threshold. Let us check how many pairs exceed the higher threshold:

```{r relaxed-relatedness}
# The higher cutoff is a subset of the pairs already reported.
rel2 <- rel[rel$KINSHIP > 2^-3.5, , drop = FALSE]
str(rel2)
```
At the higher threshold, `r nrow(rel2)` pairs are flagged. This result is calculated from the current run.

<br>

#### Principal Component Analysis (PCA)

We then exclude every individual in a flagged pair from fitting the PCs and compute PCA on the remaining individuals. This conservative teaching choice removes both members of each pair; retaining one member per family is another option. The function `bed_autoSVD()` filters MAC < 10 or MAF < 0.02, clumps variants at r² > 0.2, and iteratively removes variants in long-range LD regions. These steps reduce the risk of PCs capturing LD.

```{r initial-pca}
(obj.bed <- bed(bedfile))
stopifnot(!anyDuplicated(obj.bed$fam$sample.ID))
ind.rel <- unique(match(c(rel$IID1, rel$IID2), obj.bed$fam$sample.ID))
stopifnot(!anyNA(ind.rel))
ind.norel <- setdiff(rows_along(obj.bed), ind.rel)

set.seed(2026)
obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.norel, k = 20,
                       ncores = ncores)
```

##### Q4: How many individuals were related?

```{r}
length(unique(ind.rel))
```


##### Q5: What did we do with them?

We excluded all `r length(ind.rel)` distinct individuals in flagged pairs from fitting the PCs. We project them back later; they are not discarded from the final output.


##### Q6: How many PCs do we estimate?

20.

##### Q7: Plot the proportion of variance explained by each PC. Which PC explains most genetic variation?

The singular values are `d`; their squares measure variation along the corresponding PCs. For the standardized genotype matrix Z used in this PCA,

$$
\mathrm{PVE}_j = d_j^2 / \lVert Z \rVert_F^2. \tag{1}
$$

We calculate the denominator over the same individuals and retained variants, in small blocks. Missing genotypes contribute zero after centering (mean imputation).

```{r variance-explained, fig.cap="Figure 2. Fraction of total standardized genotype variance explained by each computed PC."}
ind.col.pca <- attr(obj.svd, "subset")
total_ss <- 0
for (first in seq.int(1L, length(ind.col.pca), by = 1024L)) {
  j <- first:min(first + 1023L, length(ind.col.pca))
  Z <- obj.bed[ind.norel, ind.col.pca[j], drop = FALSE]
  Z <- scale(Z, center = obj.svd$center[j], scale = obj.svd$scale[j])
  Z[is.na(Z)] <- 0
  total_ss <- total_ss + sum(Z^2)
}
rm(Z)
pve <- obj.svd$d^2 / total_ss
plot(pve, xlab = "Principal component", ylab = "Fraction of total variance",
     type = "b", pch = 19, col = "blue")
```

PC1 explains the most: `r round(100 * pve[1], 2)`% of total variance in this matrix.

##### Q8: Make a cumulative plot of the variance explained by the PCs. Do the first 20 PCs explain 90% of the total variation? If not, can this truncated analysis tell you exactly how many PCs are needed?

```{r cumulative-variance, fig.cap="Figure 3. Cumulative variance explained, using the full variance denominator."}
cumulative_pve <- cumsum(pve)
plot(cumulative_pve, xlab = "Number of principal components",
     ylab = "Cumulative fraction of total variance", ylim = c(0, 1),
     type = "b", pch = 19, col = "red")
abline(h = 0.9, col = "darkgreen", lty = 2)
```

The first 20 PCs explain `r round(100 * sum(pve), 2)`% of total variance.
`r if (sum(pve) < 0.9) "They do not reach 90%, so more than 20 PCs are needed; the exact number cannot be determined from these 20 singular values." else paste("The first", which(cumulative_pve >= 0.9)[1], "PCs reach 90%.")`
Normalizing `d^2` by `sum(d^2)` would instead force the cumulative sum of these 20 PCs to 100% and answer a different question.

##### Q9: Plot the PCs pairwise.

You can color the plots based on the ancestral information from the .fam2 file in the data folder.

```{r initial-score-plot, fig.width=10, fig.height=8, fig.cap="Figure 4. Initial PC pairs, colored by 1000 Genomes super-population labels."}
anc_info <- read.delim('data/1000G_phase3_common_norel.fam2')
stopifnot(!anyDuplicated(anc_info$sample.ID))
anc_info <- anc_info[match(obj.bed$fam$sample.ID, anc_info$sample.ID), ]
stopifnot(identical(anc_info$sample.ID, obj.bed$fam$sample.ID))
anc_fit <- anc_info[ind.norel, ]

cowplot::plot_grid(plotlist = lapply(1:4, function(k) {
  plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
  aes(color=anc_fit$Super.Population) +
  scale_color_viridis_d(name="Super-population")
}), scale = 0.95)
```

##### Q10: Can you describe what PCs 3-8 are capturing?

Inspect Figure 4 and the population labels. Earlier PCs separate broad ancestry patterns; later PCs may distinguish finer population structure, admixture, or outliers. A PC is not an ancestry category, and axes 3–8 should be interpreted from these plots rather than assigned a universal biological meaning.

<br>

#### Outlier sample detection (quality control)

We now look for sample outliers in PC space. They may reflect genotyping issues or unusual ancestry; an outlier score alone cannot establish poor genotype quality. We retain the 2025 exercise threshold of 0.5 for this demonstration.

```{r outlier-scores, fig.cap="Figure 5. Sample outlier statistic; the red line marks the teaching threshold of 0.5."}
prob <- bigutilsr::prob_dist(obj.svd$u, ncores = ncores)
S <- prob$dist.self / sqrt(prob$dist.nn)
stopifnot(all(is.finite(S)))

ggplot() +
  geom_histogram(aes(S), bins = 40, color = "#000000", fill = "#000000", alpha = 0.5) +
  geom_vline(xintercept = 0.5, color = "red", linetype = 2) +
  scale_y_sqrt(breaks = c(10, 100, 500)) +
  theme_bigstatsr() +
  labs(x = "Statistic of outlierness", y = "Frequency (sqrt-scale)")

```

We can then rerun the PCA without these outliers

```{r refit-pca}
ind.row <- ind.norel[S < 0.5]
ind.col <- attr(obj.svd, "subset") # starting with the previous set of variants kept (we can therefore skip the initial clumping step)
set.seed(2027)
obj.svd2 <- bed_autoSVD(obj.bed, ind.row = ind.row,
                        ind.col = ind.col, thr.r2 = NA,
                        k = 20, ncores = ncores)
```

##### Q11: How many individuals were removed?  

```{r}
sum(S >= 0.5)
```


##### Q12: What are we filtering on? (may have to read the paper to answer this) 

The calculation uses five nearest neighbours in the left singular vectors U. Let a_i be the root mean squared distance from sample i to those neighbours, and b_i the mean of their own a-values. The statistic used here is

$$ S_i = a_i / \sqrt{b_i}. \tag{2} $$

We exclude samples with S_i ≥ 0.5 from the reference fit. This is a heuristic threshold in this representation, not a calibrated probability or proof of a genotyping error. The upstream vignette currently uses 0.6; this exercise preserves its original 0.5.

<br>

#### Final PCA with all individuals

We would still like to obtain the PCA values for the remaining individuals. We do this by projecting the excluded individuals onto the newly derived PCs, to get PCs for all individuals.

```{r projection}
PCs <- matrix(NA_real_, nrow(obj.bed), ncol(obj.svd2$u))
PCs[ind.row, ] <- predict(obj.svd2)
ind.proj <- setdiff(rows_along(obj.bed), ind.row)

if (length(ind.proj)) {
  proj <- bed_projectSelfPCA(obj.svd2, obj.bed,
                            ind.row = ind.proj, ncores = 1)
  PCs[ind.proj, ] <- proj$OADP_proj
}
stopifnot(all(is.finite(PCs)))
```

##### Q13: How do we project individuals onto a PC space?  (might have to read paper or do math)
For simple projection, center and scale the held-out genotypes using the *reference* means and scales, then multiply by the reference loadings:

$$ T_{\mathrm{new}} = Z_{\mathrm{new}} V_{\mathrm{ref}}. \tag{3} $$

This can shrink projected scores. [`bed_projectSelfPCA()`](https://privefl.github.io/bigsnpr/reference/bed_projectSelfPCA.html) also returns OADP (Online Augmentation, Decomposition, and Procrustes) scores, which we use above to reduce that shrinkage.

```{r final-score-plot, fig.width=10, fig.height=8, fig.cap="Figure 6. Final PCs for every individual, including the projected individuals."}
PCs_df <- as.data.frame(PCs)
colnames(PCs_df) <- paste0("PC", seq_len(ncol(PCs)))
PCs_df$Super.Population <- anc_info$Super.Population
cowplot::plot_grid(plotlist = lapply(1:4, function(k) {
  ggplot(PCs_df, aes(x = .data[[paste0("PC", 2*k-1)]],
                    y = .data[[paste0("PC", 2*k)]],
                    color = Super.Population)) +
    geom_point(size = 0.7, alpha = 0.7) +
    scale_color_viridis_d(name = "Super-population") +
    theme_bw()
}))
```

##### Q14: Did the PCs 1–8 change? Why or why not?

Compare Figures 4 and 6, and compare scores on the individuals used in both fits. Signs may reverse without changing the PCA; near-tied components may also rotate. The second fit can change the retained variants as well as the samples.

```{r compare-pcs}
pc_cor <- abs(diag(cor(predict(obj.svd)[match(ind.row, ind.norel), 1:8],
                       predict(obj.svd2)[, 1:8])))
knitr::kable(data.frame(PC = 1:8, absolute_correlation = pc_cor), digits = 4,
             caption = "Table 1. Agreement of initial and refitted PC scores on shared reference individuals.")
```

This run excludes `r sum(S >= 0.5)` sample(s) at the outlier step and projects `r length(ind.proj)` individuals in total. Any change must be assessed from the computed comparison, rather than assuming that one excluded sample has no effect.
